# Debugging and Testing

Debugging is the process of reducing a failure to a reproducible cause. Testing
turns the expected behavior into checks that detect future regressions.

## Conditions in R

R communicates problems through conditions:

- A **message** reports progress or context.
- A **warning** signals a concern while allowing evaluation to continue.
- An **error** stops the current evaluation.

A function should fail early when continuing would produce a misleading
result.

```{r explicit-error}
safe_log <- function(x) {
  if (!is.numeric(x) || any(!is.finite(x)) || any(x <= 0)) {
    stop("x must contain positive finite numbers.")
  }
  log(x)
}

safe_log(c(1, exp(1)))
```

## Capture expected failures

`tryCatch()` handles a condition deliberately. Do not use it merely to hide an
unexpected error.

```{r try-catch}
captured_message <- tryCatch(
  safe_log(c(1, -1)),
  error = function(condition) conditionMessage(condition)
)

captured_message
```

The handler converts this intentionally demonstrated error into a value, so the
book can test the expected message without globally allowing failures.

## A reproducible debugging workflow

1. Record the exact error and call that triggered it.
2. Reduce the input to the smallest example that still fails.
3. Inspect structure with `str()`, `class()`, `typeof()`, `length()`, and
   `names()`.
4. Check assumptions at the boundary where the data enter the function.
5. Fix the cause and add a test that would have caught it.

```{r inspect-bug-input}
unexpected_input <- data.frame(
  amount = c("10", "20", "unknown")
)

str(unexpected_input)
```

The structure reveals that `amount` is character. Converting blindly would
turn the unexpected label into a missing value, so validate allowed text first.

```{r validate-conversion}
parse_amount <- function(x) {
  valid_number <- grepl("^[0-9]+([.][0-9]+)?$", x)
  if (any(!valid_number)) {
    stop("amount contains a non-numeric label.")
  }
  as.numeric(x)
}

parse_amount(c("10", "20.5"))
```

## Interactive tools

Use these tools during an interactive debugging session:

- `traceback()` displays the call stack after an error.
- `browser()` pauses inside a function.
- `debugonce(function_name)` opens the debugger on the next call.
- `recover()` lets you inspect active call frames after an error.

```{r interactive-debugging, eval=FALSE}
options(error = recover)
debugonce(model_function)
model_function(data)
options(error = NULL)
```

Do not leave global debugging options enabled in reproducible scripts.

## Lightweight assertions

`stopifnot()` is useful for internal invariants and compact checks.

```{r assertions}
normalize_weights <- function(weights) {
  stopifnot(
    is.numeric(weights),
    length(weights) > 0L,
    all(is.finite(weights)),
    all(weights >= 0),
    sum(weights) > 0
  )

  weights / sum(weights)
}

normalized_weights <- normalize_weights(c(1, 2, 1))
stopifnot(abs(sum(normalized_weights) - 1) < 1e-12)
normalized_weights
```

For a larger package or application, use a testing framework such as
`testthat` so tests are organized, named, and run automatically.

## Test boundaries and edge cases

At minimum, test:

- typical valid input;
- empty or length-zero input;
- missing and non-finite values;
- invalid types and out-of-range values;
- boundary values where conditions change;
- properties that must always hold, such as normalized weights summing to one.

## Interview checkpoints

- Distinguish messages, warnings, and errors.
- Explain when `tryCatch()` is appropriate.
- Describe how you reduce a failure to a minimal reproducible example.
- Compare input validation with internal assertions.
- Name important edge cases for a data-processing function.
