# R Foundations

R is a programming language designed for statistical computing and data
analysis. Its most important practical idea is vectorization: many operations
work on an entire vector rather than requiring an explicit loop.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  error = FALSE,
  warning = TRUE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
set.seed(2026)
```

## Objects and assignment

Use `<-` for assignment in book and analysis code. Names should describe the
meaning of an object rather than its storage type.

```{r objects-assignment}
sample_size <- 100L
conversion_rate <- 0.18
campaign_name <- "Summer launch"
is_complete <- FALSE

sample_size
```

The `L` suffix creates an integer. A number such as `0.18` is stored as a
double. Use `typeof()` for the underlying storage type and `class()` for the
object-oriented class used by generic functions.

```{r inspect-objects}
typeof(sample_size)
typeof(conversion_rate)
class(campaign_name)
class(is_complete)
```

## Atomic vectors

An atomic vector contains values of one underlying type. Arithmetic and most
base R functions operate element by element.

```{r vectorization}
revenue <- c(120, 150, 135, 180)
revenue * 1.05
mean(revenue)
```

When values of different types are combined, R coerces them to a common type.
This can be useful, but accidental coercion is a common source of bugs.

```{r coercion}
mixed <- c(10, TRUE, "20")
mixed
typeof(mixed)
```

Here every value becomes character because character is the most flexible of
the supplied types. Convert deliberately with functions such as
`as.integer()`, `as.double()`, and `as.character()`.

## Missing values

`NA` represents a missing value. Most summary functions propagate missingness
unless the analyst explicitly chooses to remove missing values.

```{r missing-values}
scores <- c(82, 91, NA, 88)

mean(scores)
mean(scores, na.rm = TRUE)
is.na(scores)
```

Do not test missingness with `x == NA`; the result is unknown rather than
`TRUE` or `FALSE`. Use `is.na(x)`.

## Comparisons and logical operators

Use `&` and `|` for element-wise comparisons. Use `&&` and `||` only when a
single Boolean decision is intended, such as the condition in an `if`
statement.

```{r logical-operators}
values <- 1:6
values > 2 & values < 6
values[values > 2 & values < 6]
```

`%in%` is useful for membership checks and never returns `NA` for ordinary
missing-value comparisons.

```{r membership}
requested_ids <- c(2, 4, 8)
values %in% requested_ids
```

## Functions and help

Functions are objects. A function receives arguments, performs a focused task,
and returns its final evaluated expression unless `return()` is used.

```{r simple-function}
center <- function(x, na.rm = FALSE) {
  x - mean(x, na.rm = na.rm)
}

center(c(2, 4, 6))
```

Inspect a function's interface with `args()` and open its documentation with
`help()` during an interactive session.

```{r function-interface}
args(mean)
```

## Paths and temporary files

Prefer project-relative paths. Avoid embedding a personal absolute path or
calling `setwd()` inside reusable analysis code.

```{r temporary-file}
temporary_path <- tempfile(fileext = ".txt")
writeLines(c("first row", "second row"), temporary_path)
readLines(temporary_path)
unlink(temporary_path)
```

`tempfile()` makes the example portable and prevents it from leaving a file in
the repository.

## Interview checkpoints

- Explain the difference between `class()` and `typeof()`.
- Describe vector coercion and give an example of accidental coercion.
- Explain why `is.na(x)` is required instead of `x == NA`.
- Distinguish `&` from `&&` and `|` from `||`.
- Explain why project-relative paths are safer than `setwd()`.
