# Iteration and Grouped Operations

Iteration means applying a computation repeatedly. In R, the best choice may
be vectorized arithmetic, a clear loop, or a member of the apply family. Choose
the form that makes the input and output structure explicit.

## Start with vectorization

Vectorized functions express many common operations directly.

```{r iteration-vectorization}
values_to_scale <- c(2, 5, 8, 11)
(values_to_scale - mean(values_to_scale)) / sd(values_to_scale)
```

Vectorization is not a rule against loops. Use a loop when it communicates the
algorithm more clearly or each iteration depends on previous state.

## `lapply()` and `vapply()`

`lapply()` always returns a list. It works naturally with data frames because a
data frame is a list of equal-length columns.

```{r lapply-example}
sample_data <- data.frame(
  units = c(5, 8, 6),
  price = c(12.5, 11.0, 13.0),
  region = c("north", "south", "west")
)

lapply(sample_data, class)
```

`vapply()` requires an output template. It fails early if a result has an
unexpected type or length.

```{r vapply-example}
vapply(sample_data, typeof, character(1))
```

`sapply()` attempts to simplify automatically. It is convenient interactively,
but `vapply()` is safer when downstream code requires a stable contract.

## Iterating over several inputs

`Map()` applies a function across inputs in parallel and returns a list.
`mapply()` uses similar logic and attempts to simplify.

```{r map-example}
means <- c(0, 5, 10)
standard_deviations <- c(1, 2, 3)

Map(
  function(mu, sigma) c(lower = mu - sigma, upper = mu + sigma),
  means,
  standard_deviations
)
```

## Matrices and margins

`apply(x, margin, fun)` operates over matrix or array margins. A margin of `1`
means rows and `2` means columns.

```{r matrix-apply}
measurement_matrix <- matrix(1:12, nrow = 3, byrow = TRUE)

apply(measurement_matrix, 1, max)
apply(measurement_matrix, 2, mean)
```

Prefer specialized functions such as `rowSums()`, `rowMeans()`, `colSums()`,
and `colMeans()` when they express the calculation directly.

```{r matrix-specialized}
rowSums(measurement_matrix)
colMeans(measurement_matrix)
```

## Split–apply–combine

Many data-analysis tasks split observations into groups, apply a summary, and
combine the results.

```{r split-apply-combine}
transactions <- data.frame(
  segment = c("new", "new", "returning", "returning"),
  revenue = c(25, 35, 60, 80)
)

by_segment <- split(transactions$revenue, transactions$segment)
vapply(by_segment, mean, numeric(1))
```

`tapply()` provides a compact form for a vector grouped by one or more factors.

```{r tapply-example}
tapply(transactions$revenue, transactions$segment, mean)
```

`aggregate()` returns a data frame and is convenient when several grouping
columns must remain explicit.

```{r aggregate-example}
aggregate(revenue ~ segment, data = transactions, FUN = mean)
```

## Stateful iteration

When the next value depends on the previous value, a loop is usually clearer
than forcing the problem into an apply function.

```{r stateful-iteration}
growth_rates <- c(0.10, -0.05, 0.20)
balance_path <- numeric(length(growth_rates) + 1L)
balance_path[1] <- 100

for (i in seq_along(growth_rates)) {
  balance_path[i + 1L] <- balance_path[i] * (1 + growth_rates[i])
}

round(balance_path, 2)
```

## Interview checkpoints

- Compare `lapply()`, `sapply()`, and `vapply()`.
- Explain what the `MARGIN` argument means in `apply()`.
- Describe split–apply–combine and name two R implementations.
- Explain when a `for` loop is clearer than an apply-family function.
- Explain why `rowMeans()` is preferable to `apply(x, 1, mean)` when applicable.
