# Control Flow and Functions

Control flow determines which expressions run and how often they run.
Functions package those decisions into reusable units that can be tested and
combined.

## Scalar decisions

An `if` condition must represent one logical decision. Test missingness before
using a value in a condition because `if (NA)` cannot choose a branch.

```{r scalar-condition}
classify_score <- function(score) {
  if (is.na(score)) {
    "missing"
  } else if (score >= 90) {
    "high"
  } else if (score >= 70) {
    "medium"
  } else {
    "low"
  }
}

classify_score(84)
classify_score(NA_real_)
```

Use `&&` and `||` for scalar control-flow conditions. They evaluate from left
to right and stop as soon as the result is known.

```{r short-circuiting}
is_valid_rate <- function(x) {
  length(x) == 1L && !is.na(x) && x >= 0 && x <= 1
}

is_valid_rate(0.4)
is_valid_rate(NA_real_)
```

## Vectorized decisions

`ifelse()` chooses element by element. It is convenient for simple recoding,
but the output type is determined by its branches and may not preserve every
class attribute.

```{r vectorized-condition}
response_times <- c(1.2, 2.8, NA, 4.1)
status <- ifelse(
  is.na(response_times),
  "missing",
  ifelse(response_times <= 3, "on time", "late")
)
status
```

For complex business rules, write explicit tests and verify that categories
are mutually exclusive and collectively exhaustive.

## `for` loops

A loop is appropriate when each step has side effects or depends on previous
steps. Preallocate the result instead of repeatedly growing an object.

```{r preallocated-loop}
inputs <- c(2, 4, 6, 8)
squared <- numeric(length(inputs))

for (i in seq_along(inputs)) {
  squared[i] <- inputs[i]^2
}

squared
```

`seq_along(x)` safely handles an empty vector, unlike `1:length(x)`.

## `while` loops

A `while` loop is useful when the number of iterations is not known in advance.
Its state must move toward a stopping condition.

```{r while-loop}
remaining <- 100
periods <- 0L

while (remaining > 10) {
  remaining <- remaining * 0.6
  periods <- periods + 1L
}

c(periods = periods, remaining = remaining)
```

In production code, consider a maximum-iteration guard when convergence is not
guaranteed.

## Function contracts

A useful function has a focused responsibility, meaningful argument names,
and clear behavior for invalid input.

```{r function-contract}
weighted_mean <- function(x, weights, na.rm = FALSE) {
  if (!is.numeric(x) || !is.numeric(weights)) {
    stop("x and weights must be numeric.")
  }
  if (length(x) != length(weights)) {
    stop("x and weights must have equal lengths.")
  }
  if (any(weights < 0, na.rm = TRUE)) {
    stop("weights must be non-negative.")
  }

  stats::weighted.mean(x, weights, na.rm = na.rm)
}

weighted_mean(c(10, 20, 30), c(1, 2, 1))
```

Defaults should be safe and unsurprising. Validate assumptions close to the
function boundary so failures explain what is wrong.

## Lexical scope and closures

R resolves names in the environment where a function was created. A closure is
a function bundled with that environment.

```{r closure-example}
make_counter <- function(start = 0L) {
  value <- as.integer(start)

  function() {
    value <<- value + 1L
    value
  }
}

counter <- make_counter(10)
counter()
counter()
```

`<<-` changes a binding outside the current function. It can support deliberate
state, caching, or counters, but ordinary analysis functions should usually
return a value instead of modifying hidden state.

## Pure functions

A pure function returns the same result for the same inputs and does not alter
external state. Pure functions are easier to test and reuse.

```{r pure-function}
standardize <- function(x) {
  (x - mean(x)) / stats::sd(x)
}

round(standardize(c(10, 12, 14, 16)), 3)
```

## Interview checkpoints

- Distinguish scalar `if` from vectorized `ifelse()`.
- Explain short-circuit evaluation with `&&` and `||`.
- Explain why preallocation matters in a loop.
- Describe lexical scope, closures, and the risks of `<<-`.
- Define a pure function and explain why it is easier to test.
