# Simulation and Reproducibility

Simulation helps test intuition, evaluate estimators, and understand sampling
variation. A reproducible simulation specifies the random seed, data-generating
process, number of repetitions, and quantity being estimated.

## Distribution functions

R uses consistent prefixes for probability distributions:

- `d*` returns a density or probability mass.
- `p*` returns a cumulative probability.
- `q*` returns a quantile.
- `r*` generates random values.

```{r distribution-functions}
dnorm(0)
pnorm(1.96)
qnorm(0.975)
```

For a continuous distribution, `dnorm(0)` is a density height, not the
probability that a continuous random variable equals exactly zero.

## Reproducible random generation

`set.seed()` resets R's random-number stream so the same code produces the same
sequence.

```{r reproducible-randomness}
set.seed(2026)
first_draw <- rnorm(5)

set.seed(2026)
second_draw <- rnorm(5)

identical(first_draw, second_draw)
```

Set a seed at the simulation boundary. Repeatedly resetting it inside a helper
function can accidentally generate identical samples across calls.

## Sampling

Sampling without replacement selects distinct values. Sampling with
replacement permits a value to appear multiple times and underlies the
nonparametric bootstrap.

```{r sampling}
set.seed(2026)
sample(1:10, size = 4, replace = FALSE)
sample(1:6, size = 8, replace = TRUE)
```

## Simulating a linear model

Make the data-generating process explicit before fitting the model.

```{r linear-model-simulation}
set.seed(2026)
simulation_size <- 500
predictor <- rnorm(simulation_size)
noise <- rnorm(simulation_size, mean = 0, sd = 2)
outcome <- 0.5 + 2 * predictor + noise

simulated_fit <- lm(outcome ~ predictor)
round(coef(simulated_fit), 2)
```

The fitted coefficients will not equal the generating parameters exactly
because each sample contains random variation.

## Monte Carlo estimation

A Monte Carlo estimate is an average over repeated random experiments. Its
simulation error generally decreases as the number of repetitions increases.

```{r monte-carlo}
set.seed(2026)
repetitions <- 10000
dice_totals <- replicate(
  repetitions,
  sum(sample(1:6, size = 2, replace = TRUE))
)

mean(dice_totals >= 10)
```

Report the seed and repetition count so another analyst can reproduce the
result and assess simulation precision.

## Bootstrap uncertainty

The nonparametric bootstrap repeatedly samples observed values with replacement
and recomputes a statistic.

```{r bootstrap-example}
observed_revenue <- c(20, 25, 30, 35, 50, 70)

set.seed(2026)
bootstrap_means <- replicate(
  2000,
  mean(sample(observed_revenue, replace = TRUE))
)

round(quantile(bootstrap_means, c(0.025, 0.975)), 2)
```

This interval reflects the empirical sample and bootstrap assumptions. It is
not a substitute for checking whether observations are independent and
representative of the target population.

## Timing and profiling

Optimize only after identifying a real bottleneck. `system.time()` measures an
expression, while `Rprof()` and dedicated profilers show where a program spends
time.

```{r profiling-example, eval=FALSE}
system.time({
  result <- expensive_function(data)
})

Rprof("profile.out")
result <- expensive_function(data)
Rprof(NULL)
summaryRprof("profile.out")
```

Profiling examples are not executed here because they require a real workload
and create output files.

## Interview checkpoints

- Explain the `d`, `p`, `q`, and `r` distribution prefixes.
- Explain what `set.seed()` guarantees and what it does not.
- Distinguish sampling with and without replacement.
- Describe a data-generating process for a simple linear model.
- Explain what uncertainty a nonparametric bootstrap represents.
