# Data Manipulation

Data manipulation should make row selection, variable creation, aggregation,
and table relationships visible. Start by checking keys, types, and row counts;
then verify that each transformation preserves the intended unit of analysis.

## Example tables

```{r manipulation-data}
customers_table <- data.frame(
  customer_id = 1:4,
  segment = c("new", "returning", "new", "returning")
)

orders_table <- data.frame(
  order_id = 101:106,
  customer_id = c(1, 2, 2, 3, 4, 4),
  revenue = c(30, 45, 55, 25, 60, 40)
)

customers_table
orders_table
```

`customer_id` is unique in the customer table but repeated in the order table.
That makes a join from orders to customers many-to-one.

## Base R selection and transformation

Rows can be filtered with a logical vector and columns selected by name.

```{r base-selection}
orders_table[
  orders_table$revenue >= 40,
  c("order_id", "customer_id", "revenue")
]
```

Assign a derived column directly when the formula is simple and vectorized.

```{r base-transformation}
orders_table$revenue_after_fee <- orders_table$revenue * 0.97
head(orders_table, 3)
```

## `dplyr` pipelines

The `dplyr` verbs make a sequence of table operations explicit. Use namespace
qualification in reusable examples so readers can see where a function comes
from.

```{r dplyr-pipeline}
high_value_orders <- orders_table |>
  dplyr::filter(revenue >= 40) |>
  dplyr::mutate(
    revenue_band = dplyr::if_else(revenue >= 55, "high", "medium")
  ) |>
  dplyr::select(order_id, customer_id, revenue, revenue_band) |>
  dplyr::arrange(dplyr::desc(revenue))

high_value_orders
```

Inside a `dplyr` data mask, `if_else()` is type-stable and requires compatible
true, false, and missing branches.

## Grouped summaries

After grouping, every summary should have a clear interpretation at the new
unit of analysis.

```{r grouped-summary}
customer_summary <- orders_table |>
  dplyr::group_by(customer_id) |>
  dplyr::summarise(
    order_count = dplyr::n(),
    total_revenue = sum(revenue),
    average_revenue = mean(revenue),
    .groups = "drop"
  )

customer_summary
```

Always check whether missing values should propagate or be explicitly removed.

## Joins and cardinality

State the expected relationship before joining. A many-to-many join can
multiply rows and totals even when the code runs without an error.

```{r validated-join}
orders_enriched <- dplyr::left_join(
  orders_table,
  customers_table,
  by = "customer_id",
  relationship = "many-to-one"
)

orders_enriched
```

Reconcile important quantities after the join.

```{r join-reconciliation}
stopifnot(
  nrow(orders_enriched) == nrow(orders_table),
  sum(orders_enriched$revenue) == sum(orders_table$revenue),
  !anyNA(orders_enriched$segment)
)
```

## Avoid row-wise work when possible

Vectorized expressions and grouped operations are usually clearer than
processing one row at a time. Row-wise logic is appropriate only when the task
truly operates on heterogeneous values within each record.

```{r vectorized-derived-column}
orders_enriched$revenue_share <-
  orders_enriched$revenue / sum(orders_enriched$revenue)

round(sum(orders_enriched$revenue_share), 10)
```

## Interview checkpoints

- Define a table's unit of analysis and candidate key.
- Explain one-to-one, one-to-many, and many-to-many joins.
- Describe how a join can inflate row counts and aggregate totals.
- Explain the purpose of `.groups = "drop"` after `summarise()`.
- Describe checks you would perform before and after a production join.
