# Missing and Non-finite Values

Missingness is part of the data-generating process, not merely a cleaning
inconvenience. Before removing or imputing values, ask why they are absent and
whether that absence contains information.

## `NA`, `NaN`, and infinity

R distinguishes missing values from non-finite numerical results.

```{r nonfinite-values}
special_values <- c(10, NA_real_, NaN, Inf, -Inf)

is.na(special_values)
is.nan(special_values)
is.infinite(special_values)
is.finite(special_values)
```

`is.na()` is also `TRUE` for `NaN`, while `is.nan()` identifies only undefined
numeric results such as `0 / 0`.

## Missingness propagates

R does not silently discard missing values from most summaries.

```{r missing-propagation}
observed_scores <- c(72, 85, NA, 91)

mean(observed_scores)
mean(observed_scores, na.rm = TRUE)
```

Using `na.rm = TRUE` changes the population being summarized. Make that choice
deliberately and report how many observations were excluded.

```{r missing-counts}
c(
  total = length(observed_scores),
  missing = sum(is.na(observed_scores)),
  observed = sum(!is.na(observed_scores))
)
```

## Row completeness

`complete.cases()` identifies rows without missing values across selected
variables.

```{r complete-cases}
applications <- data.frame(
  applicant_id = 1:5,
  income = c(60, 75, NA, 90, 68),
  debt = c(20, NA, 25, 30, 18),
  approved = c(TRUE, FALSE, TRUE, TRUE, FALSE)
)

analysis_columns <- c("income", "debt", "approved")
complete_rows <- complete.cases(applications[analysis_columns])

applications[complete_rows, ]
```

Complete-case analysis is valid only under assumptions about why values are
missing. It can reduce precision and introduce bias when missingness is related
to the outcome or predictors.

## Missingness mechanisms

Three common conceptual categories are:

- **MCAR:** missingness is unrelated to observed or unobserved values.
- **MAR:** after conditioning on observed information, missingness does not
  depend on the missing value itself.
- **MNAR:** missingness still depends on unobserved information or the missing
  value after conditioning on observed data.

These mechanisms are assumptions about the process, not labels that can
usually be proven from the dataset alone.

## Simple imputation

Median imputation can be a reasonable baseline for a numeric predictor, but it
reduces variation and does not automatically remove bias.

```{r median-imputation}
training_income <- c(55, 62, 70, NA, 88)
training_median <- median(training_income, na.rm = TRUE)

impute_median <- function(x, value) {
  x[is.na(x)] <- value
  x
}

impute_median(training_income, training_median)
```

For predictive modeling, estimate the imputation value using the training data
only, then apply that saved value to validation, test, and future data. Fitting
preprocessing on the full dataset leaks information across the evaluation
boundary.

## Missing indicators

Sometimes the fact that a value is missing is predictive. A missingness
indicator can preserve that signal, but it does not eliminate the need to
understand the missingness process.

```{r missing-indicator}
applications$income_missing <- is.na(applications$income)
applications$income_imputed <- impute_median(
  applications$income,
  median(applications$income, na.rm = TRUE)
)

applications[c("applicant_id", "income_missing", "income_imputed")]
```

## Safe comparisons

Comparing a missing value with `==` produces `NA`, not a Boolean answer.

```{r safe-missing-comparison}
applications$income == NA
is.na(applications$income)
```

Use `is.na()` for missingness and `is.finite()` when a calculation requires an
ordinary finite number.

## Interview checkpoints

- Distinguish `NA`, `NaN`, and `Inf`.
- Explain why `x == NA` does not test missingness.
- Describe MCAR, MAR, and MNAR as assumptions.
- Explain how preprocessing can leak information from validation or test data.
- Discuss one benefit and one limitation of a missingness indicator.
