# Vectors and Data Structures

Choosing the correct data structure makes R code easier to reason about. The
core structures differ by dimensionality and whether they permit mixed types.

| Structure | Dimensions | Value types |
| --- | ---: | --- |
| Atomic vector | 1 | One underlying type |
| Matrix | 2 | One underlying type |
| List | 1 | Mixed objects allowed |
| Data frame | 2 | Each column may have a different type |

## Creating sequences

Use `seq()` when the start, end, or increment should be explicit. Use
`seq_along(x)` when indices must follow the length of another object.

```{r sequences}
seq(from = 2, to = 10, by = 2)
seq_len(5)

labels <- c("A", "B", "C")
seq_along(labels)
```

`seq_along()` is safer than `1:length(x)` because it correctly returns an empty
integer vector when `x` has length zero.

## Indexing vectors

R supports position, logical, and name-based indexing.

```{r vector-indexing}
sales <- c(north = 120, south = 95, west = 140, east = 110)

sales[c(1, 3)]
sales[sales >= 120]
sales[c("north", "east")]
```

Negative positions exclude elements. Do not mix positive and negative
positions in the same index.

```{r negative-indexing}
sales[-2]
```

## Matrices

A matrix is an atomic vector with a two-dimensional `dim` attribute. Because a
matrix has one underlying type, inserting character data coerces numeric values
to character.

```{r matrices}
measurements <- matrix(
  1:12,
  nrow = 3,
  byrow = TRUE,
  dimnames = list(NULL, c("a", "b", "c", "d"))
)

measurements
measurements[, c("a", "c"), drop = FALSE]
```

`drop = FALSE` preserves the matrix structure when a selection contains only
one row or column.

## Lists

A list can contain objects of different types and sizes.

```{r lists}
model_summary <- list(
  model = "linear regression",
  coefficients = c(intercept = 1.2, slope = 0.8),
  converged = TRUE
)

model_summary[["coefficients"]]
model_summary["coefficients"]
```

`[[` extracts the object stored in one list element. `[` returns a sub-list.
That distinction is a frequent interview question and a common source of
unexpected results.

## Data frames

A data frame is a list of equal-length columns. Columns may use different
types, but each column must have a consistent underlying type.

```{r data-frames}
customers <- data.frame(
  customer_id = 101:104,
  segment = c("new", "returning", "new", "returning"),
  revenue = c(35.0, 82.5, 41.0, NA_real_)
)

str(customers)
customers[customers$segment == "new", c("customer_id", "revenue")]
```

Use `NA_real_`, `NA_integer_`, and related typed missing values when the desired
column type should be explicit.

## Factors

A factor stores categorical values as integer codes with associated labels.
Use an ordered factor only when the levels have a meaningful ranking.

```{r factors}
priority <- factor(
  c("medium", "high", "low", "medium"),
  levels = c("low", "medium", "high"),
  ordered = TRUE
)

priority
priority > "low"
```

Never rely on alphabetical ordering when the domain defines the order.

## Combining data

`cbind()` combines objects by columns and `rbind()` combines them by rows. R
may recycle shorter vectors, sometimes with only a warning, so validate lengths
before combining independent inputs.

```{r combining-data}
customer_id <- 1:3
revenue <- c(25, 40, 30)

data.frame(customer_id, revenue)
```

For production analysis, prefer joining tables by explicit keys rather than
assuming two independently created objects have the same row order.

## Inspecting unfamiliar objects

Use a small set of inspection functions before manipulating unfamiliar data.

```{r inspecting-data}
dim(customers)
names(customers)
head(customers, 2)
summary(customers)
```

`str()` is usually the most concise first check because it exposes dimensions,
column types, and representative values.

## Interview checkpoints

- Compare an atomic vector, matrix, list, and data frame.
- Explain the difference between `[` and `[[` for lists.
- Explain when matrix subsetting drops dimensions and how to prevent it.
- Describe how factors store values and why explicit level order matters.
- Explain why joining by keys is safer than relying on row order.
